```{r hidden-load}
#| echo: false

library("knitr")
library("ggplot2")
```

# Squashing Bugs with R's Debugging Tools

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Describe R's conditions system
* Explain what defensive programming is
* Use the conditions system to raise messages, warnings, and errors
* Use the conditions system to catch messages, warnings, and errors
* Use R's debugging functions to diagnose and fix bugs in code
* Estimate the amount of memory a data set will require
* Use the lobstr package to get memory usage for an R object
* Describe what a profiler is and why you would use one
* Describe what kinds of profiling tools R provides
:::

**Debugging** is the process of investigating why code does something
unexpected, incorrect, or erroneous. In other words, we debug software in order
to find the causes of bugs, so that we can fix the code and eliminate the bugs.

This chapter explains several strategies for debugging R code, as well as ways
to prevent or minimize bugs as you write code. It explains how to print output,
how R's conditions system for warnings and errors works, and how to use R's
debugging functions, as well as introducing testing frameworks you can use to
write tests that automatically check code for certain kinds of bugs. In some
applications, compute time and memory usage matter almost as much as bugs, so
this chapter also explains how to estimate the performance of R code.


## Know the Conditions System {#sec-the-conditions-system}

R's **conditions system** provides a way to signal and handle unusual
conditions that arise while code runs. The system mainly deals with three kinds
of conditions:

* **Messages** provide status information about running code, without signaling
  that something has gone wrong. You can use messages to print out any
  information you think might be relevant to users.
* **Warnings** signal that something unexpected happened but the code can
  continue running. By default, R prints warnings only after all entered code
  finishes running. This can make it difficult to understand the cause of a
  particular warning; @sec-global-options explains how to change this setting.
* **Errors** signal that something unexpected happened and that the code must
  immediately stop running. R prints errors as soon as they occur.

With a clear understanding of the conditions system, you can:

* Gather more information about warnings and errors, such as the line of code
  that caused them.
* **Raise** (send) conditions in your own code to signal problems. For example,
  if you write a function to read and clean a specific data format, you might
  want it to raise a warning if some columns are unexpectedly missing. This is
  especially important for code that will be reused many times or by many
  people.
* Run code that might raise conditions and handle them safely and
  appropriately. For example, if your code tries to download a file from the
  internet, you might want it to try again if the first attempt raises an error
  because the connection timed out.
* Suppress printing of warnings that you've already investigated and determined
  are not problems.

In short, the conditions system is the key to making sense of and fixing
warning and errors. It will also enable you write code that's easier to use and
more robust. We'll present several examples of how to use the conditions system
later in this chapter.


## Write Code Defensively

You might be familiar with defensive driving: the idea that when you drive a
vehicle, you should try to anticipate problems outside of your control, so that
you can react safely or avoid them entirely. Defensive driving reduces the risk
for you and others on the road.

When you write code, you can take a similar approach, called **defensive
programming**. In this case, the idea is to try to anticipate problems that
might occur as the code runs. As examples of problems you can anticipate,
suppose your code:

* Reads data from a file. What if the path to the file is invalid? What if R
  doesn't have permission to access the file? What if the structure of the data
  in the file is different from what your code assumes?
* Computes distances between field sites. What if some of the input data are
  missing or outliers? What if one of the results is negative? What if one of
  the results is larger than physically possible for the data?
* Gets the current time, to compare against times in a dataset. What if the
  dataset and the computer use different time zones? What if the dataset
  includes data from multiple time zones? What if the times have different
  resolutions (nanoseconds versus milliseconds versus seconds)?

Depending on the severity and nature of the problem, you might want the code to
print a warning message, do something special to recover, or raise an error and
stop running. The benefit of defensive programming is that it can help prevent
bugs or make them easier to detect early, so that they can be fixed swiftly.

The subsequent sections describe some functions that are helpful for
identifying problems, how to signal warnings and errors with the conditions
system, how to intercept and handle warnings and errors, and packages you can
use to add automated tests to your code.


### Check for Problems {#sec-check-for-problems}

R provides a variety of built-in functions you can use to check whether objects
have specific values, types, or structures. Many of these functions have names
that begin with `is.`, such as:

* `is.logical`, `is.integer`, `is.double`, `is.complex`, `is.character` to
  check the type of an object.
* `is.numeric` to check whether an object is a number.
* `is.atomic` to check whether a object is an atomic vector (as opposed to a
  recursive data structure, such as a list).
* `is.na`, `is.null`, `is.nan`, `is.infinite` to check for missing or special
  values.

A few more functions for checking properties of objects are:

* `inherits` to check whether an object has a specific class.
* `identical` to check whether two objects are exactly the same.
* `all.equal` to check whether two objects are approximately equal.
* Comparison operators (`<`, `<=`, `==`, and so on).

You can use these functions together with if statements to make your code react
to specific problems you anticipate. In some cases, you might want to raise a
warning or error; the next section explains how to do so.

:::{.callout-note title="See Also"}
The [rlang][] package provides alternatives with standardized interfaces for
many of these functions, as well as some additional functions.

[rlang]: https://rlang.r-lib.org/
:::


### Raise Conditions {#sec-raise-conditions}

There are three built-in functions for signaling conditions:

* `message` sends a message.
* `warning` raises a warning.
* `stop` raises an error (and, in most cases, stops R from running any further
  code).

These functions have similar but not identical parameters. First, each accepts
any number of unnamed arguments as a description of the condition. These
arguments are concatenated with no separator between them. For example:

```{r}
message("Hello, ", "how are you?")
warning("Objects in mirror", " ", "may be closer than they appear.")
```

R prints `Warning:` before the description for warnings, and `Error:` before
the description for errors. By default, calling `warning` or `stop` from the
body of a function also prints the name of the function:

```{r}
f = function(x, y) {
  warning("This is a warning!")
  x + y
}
f(3, 4)
```

The name of the function that raised the condition is generally useful
information for users that want to correct whatever caused it. Occasionally,
you might want to disable this behavior, which you can do by setting `call. =
FALSE`:

```{r}
f = function(x, y) {
  warning("This is a warning!", call. = FALSE)
  x + y
}

f(3, 4)
```

The `warning` function also has several other parameters that control when and
how warnings are displayed.

Unlike `message` and `warning`, the `stop` function immediately stops the
evaluation of code. For instance, in this function, the line `x + y` never
runs:

```{r, error = TRUE}
f = function(x, y) {
  stop("This is an error!")
  x + y
}

f(3, 4)
```

When writing code---especially functions, executable scripts, and
packages---it's a good habit to include tests for unexpected conditions such as
invalid arguments and impossible results. If the tests detect a problem, use
the `warning` or `stop` function (depending on severity) to signal what the
problem is. Try to provide a concise but descriptive warning or error message
so that users can easily understand what went wrong.


:::{.callout-note title="See Also"}
The [assertions][] package provides a concise way to check criteria and raise
warnings or errors (with custom descriptions) if they aren't met. You can
approximate this with R's built-in `stopifnot` function, but it's not as
flexible and user friendly (for example, you can't set the error description).

[assertions]: https://selkamand.github.io/assertions/
:::


### Handle Conditions

In some cases, you can anticipate the problems likely to occur when code runs
and even devise ways to work around them. As an example, suppose your code
needs to load parameters from a configuration file, but the path to the file
provided by the user is invalid. It might still be possible for your code to
run by falling back on a set of default parameters. R's conditions system
provides a way to **catch** (intercept) messages, warnings, and errors, and to
run alternative code in response.

You can use the `try` function to safely run code that might produce an error.
If no error occurs, the `try` function returns whatever the result of the code
was. If an error does occur, the `try` function prints the error message and
returns an object of class `try-error`, but evaluation does not stop. For
example:


```{r}
bad_add = function(x) {
  # No error
  x1 = try(5 + x)
  # Error
  x2 = try("yay" + x)

  list(x1, x2)
}

bad_add(10)
```

The simplest thing you can do in response to an error is ignore it. This is
usually not a good idea, but if you understand exactly what went wrong, can't
fix it easily, and know it won't affect the rest of your code, doing so might
be the best option.

A more robust approach is to inspect the result from a call to `try` to see if
an error occurred, and then take some appropriate action if one did. You can
use the `inherits` function to check whether an object has a specific class, so
here's a template for how to run code that might cause an error, check for the
error, and respond to it:

```{r, eval = FALSE}
result = try({
  # Code that might cause an error.
})
if (inherits(result, "try-error")) {
  # Code to respond to the error.
}
```

You can prevent the `try` function from printing error messages by setting
`silent = TRUE`. This is useful when your code is designed to detect and handle
the error, so you don't users to think an error occurred.

The `tryCatch` function provides another way to handle conditions raised by a
piece of code. It requires that you provide a **handler function** for each
kind of condition you want to handle. The kinds of conditions are:

* `message`
* `warning`
* `error`
* `interrupt` -- when the user interrupts the code (for example, by
  pressing Ctrl-C)

Each handler function must accept exactly one argument.

When you call `tryCatch`, if the suspect code raises a condition, then it calls
the associated handler function and returns whatever the handler returns.
Otherwise, `tryCatch` returns the result of the code.

Here's an example of using `tryCatch` to catch an error:
```{r}
bad_fn = function(x, y) {
  stop("Hi")
  x + y
}

err = tryCatch(bad_fn(3, 4), error = function(e) e)
```

And here's an example of using `tryCatch` to catch a message:

```{r}
msg_fn = function(x, y) {
  message("Hi")
  x + y
}

msg = tryCatch(msg_fn(3, 4), message = function(e) e)
```

The `tryCatch` function always silences conditions. Details about raised
conditions are provided in the object passed to the handler function, which has
class `condition` (and a more specific class that indicates what kind of
condition it is).

:::{.callout-note title="See Also"}
To learn more about R's conditions system, first read the documentation at
`?conditions`.

If you want even more details, see [this chapter][adv-r-conditions] of
*Advanced R* by Wickham.

[adv-r-conditions]: https://adv-r.hadley.nz/conditions.html
:::


### Create Automated Tests

By turning your R code into a package, you can set up a suite of automated
tests. Then you can run the tests any time you or a collaborator make changes
to the code, in order to verify that the code still works as intended. You can
also add new tests as you add new features. With a test suite, you can verify
that:

* The individual components of the package work correctly. Tests of this sort
  are called **unit tests**.
* The components of the package work together correctly. Tests of this sort are
  called **integration tests**.
* Fixed bugs stay fixed. Tests of this sort are called **regression tests**.

Automated testing is especially important for code that will be reused and
edited frequently, because each edit can introduce new bugs. It's also a good
idea to create automated tests for any code where bugs might have severe
consequences.

On the other hand, for a typical data analysis script, which you might only run
1-2 times, the effort it takes to set up automated tests usually outweighs the
benefit. Use other defensive programming techniques instead!

The remainder of this section assumes you plan to package your R code and want
to create automated tests.

:::{.callout-note}
The book [*R Packages*][r-pkgs] by Wickham and Bryan is an excellent
introduction to packaging R code. It includes multiple chapters about automated
testing.

[r-pkgs]: https://r-pkgs.org/
:::

In the directory structure for your package, R will treat any script (`.R`
file) you add to the `tests/` subdirectory as an automated test. When R runs
a test, it compares the output to a saved output file (`.Rout.save` file) of
the same name. If the output matches, the test passes. Otherwise, the test
fails. R runs all tests automatically whenever a package is built and
installed. You can also run them manually with the `R CMD check` command line
program or the `test` function from the [devtools][] package. You can read more
about R's built-in system for automated tests in the [Writing R
Extensions][r-extensions] manual.

[devtools]: https://devtools.r-lib.org/
[r-extensions]: https://cran.r-project.org/doc/manuals/r-release/R-exts.html

The built-in testing system is slightly cumbersome, since you have to prepare
an output file for every test script. Fortunately, there are a few packages
designed to make testing more convenient:

* [testthat][] is the most popular testing package. It's part of R-Lib, a
  collection of developer-oriented packages from the creators of the Tidyverse.
  As a result, it has simple, consistent interfaces and detailed documentation
  with many examples. Due to its popularity, there are also many other learning
  resources about the package available online. To write tests with testthat,
  create scripts in the `tests/testthat/` subdirectory. In the scripts, use the
  `test_that` function to register a test. Within each test, you can use the
  package's `expect_` functions to check for specific criteria. See the
  documentation or the book [*R Packages*][r-pkgs] for more details.
* [tinytest][] is a testing package designed to be simple and have minimal
  dependencies on other packages. The interface is deliberately similar to
  testthat. The package also provides functions to set up test files and run
  the tests.

[testthat]: https://testthat.r-lib.org/
[tinytest]: https://github.com/markvanderloo/tinytest



## Find Out What's Happening

If you suspect there's a bug in your code, the first thing to do is gather more
information about what the cause might be. This section provides a few
relatively simple strategies to find out what's happening without resorting to
R's debugger (which is covered later, in @sec-debug-the-code). You can also use
these strategies to help you understand the code better, even you're not trying
to fix a bug.


### Go One Step at a Time {#sec-go-one-step-at-a-time}

A crucial skill in programming (and problem solving in general) is breaking
problems down into smaller steps. Most of the time, the steps need to be
carried out in a specific order, and if one step is incorrect, it will affect
many of the steps after it. For example, you might remove outliers from a
dataset before computing a mean. If you remove the wrong values, then it's
likely the mean will also be incorrect. The point is that you must verify the
output from each step before considering the next one. In the example, you can
spend a long time trying to understand why the mean is incorrect, but until you
look at the step that removes outliers, you won't find the answer.

In terms of R code, working through a problem step by step means running the
code one expression (generally, one line) at a time. After running an
expression, take time to inspect the results. That means that if the code
changed a variable, you should look at the value of the variable. Confirm that
it matches your expectations (Is it within the range of plausible values? Can
you explain why it is what it is?). If doesn't, there might be something wrong
with the expression, or you might need rethink your understanding of what the
expression does.

:::{.callout-tip}
It's also critical to start from the beginning. That means the first expression
in your code, but it also means in a fresh R session. Restart R before you run
the code, so that there aren't any leftovers from code you ran earlier. These
leftovers can obscure bugs and produce misleading results. In some cases,
restarting R is enough to fix a bug!

If you're using RStudio, you can restart R (without closing RStudio) through
the "Session" menu.
:::


### Print the Details

Some expressions do so much that they're hard to understand even if you run the
code one step at a time (@sec-go-one-step-at-a-time). Loops are a great example
of this. One way to get information about what's happening in a loop is to make
the loop print out the details as it runs. Insert calls to the `print` function
to print out values of variables. Printing things out is a good diagnostic
strategy in general, and you can use it for any code, not just loops.

The `message` function and the `print` function both print things out, so you
might wonder why we suggest `print` here. The `message` function is intended to
print status messages to users, which is why it's part of R's conditions system
(@sec-the-conditions-system). In contrast, the `print` function is intended to
print useful information for developers. Unlike `message`, it can print
non-string objects, and it will often print extra information about the
structure of an object.

For example, when printing a vector, the `print` function prints the position
of the first element on each line in square brackets `[ ]`:

```{r}
print(1:100)
```

The `print` function also prints quotes around strings:

```{r}
print("Hi")
```

These features make the `print` function ideal for printing information when
you're trying to understand some code or diagnose a bug.


### Read Warnings and Errors {#sec-read-warnings-and-errors}

A computer does exactly what you tell it to do, even if you tell it to do
something that doesn't make sense. There are two ways this can cause problems:
if your instructions are impossible or if your instructions are incorrect. When
you tell R to do something impossible, it will typically generate an error or
warning, which will include a message that tries to tell you what went wrong.
Read these messages! Often, they describe the problem and suggest a solution.

For example:

```{r add-characters}
#| error: true
"7" + 5
```

We can't add the text `"7"` to the number `5`. To make this code work, we have
to change `"7"` to `7`.

As another example, consider the `read.csv` function, which reads a
comma-separated values file. What happens if you ask `read.csv` to read a file
that doesn't exist?

```{r file-not-exist}
#| error: true
read.csv("nope")
```

The error message doesn't tell you that the file doesn't exist, but it does say
that R was unable to open a connection to the file. Since the functions built
into R are relatively well-tested (by every R user), it makes sense to doubt
the user input before doubting R. In this case, the only input was the file
name, so a good first step is to check that the file actually exists.


### Set Global Options {#sec-global-options}

R's **global options** to control many different aspects of how R works.
They're relevant to the theme of this chapter because some of them control when
and how R displays warnings and errors.

You can use the `options` function to get or set global options. If you call
the function with no arguments, it returns the current settings:

```{r}
opts = options()
# Display the first 6 options.
head(opts)
```

This section only explains a few of the options, but you can read about all of
them in `?options`.

The `warn` option controls how R handles warnings. It can be set to three
different values:

* `0` -- (the default) warnings are only displayed after code finishes
  running.
* `1` -- warnings are displayed immediately.
* `2` -- warnings stop code from running, like errors.

Setting `warn = 2` is useful for pinpointing expressions that raise warnings.
Setting `warn = 1` makes it easier to determine which expressions raise
warnings, without the inconvenience of stopping code from running. That makes
it a good default (better than the actual default). You can use the `option`
function to change the value of the `warn` option:

```{r}
options(warn = 1)
```

When you set an option this way, the change only lasts until you quit R. Next
time you start R, the option will go back to its default value. Fortunately,
there is a way override the default options every time R starts.

When R starts, it searches for a `.Rprofile` file. The file is usually in your
system's home directory (see [this section][homedir] and Appendix A of the R
Basics Reader for how to locate your home directory). Customizing your
`.Rprofile` file is one of the marks of an experienced R user. If you define a
`.First` function in your `.Rprofile`, R will call it automatically during
startup. Here's an example `.First` function:

```{r, eval = FALSE}
.First = function() {
  # Only change options if R is running interactively.
  if (!interactive())
    return()

  options(
    # Don't print more than 1000 elements of anything.
    max.print = 1000,
    # Warn on partial matches.
    warnPartialMatchAttr = TRUE,
    warnPartialMatchDollar = TRUE,
    warnPartialMatchArgs = TRUE,
    # Print warnings immediately (2 = warnings are errors).
    warn = 1
  )
}
```

[homedir]: https://ucdavisdatalab.github.io/workshop_r_basics/chapters/getting-started/working-with-files.html#the-working-directory

:::{.callout-note title="See Also"}
You can learn more about the `.Rprofile` file and R's startup process at
`?Startup`.
:::


## Debug the Code {#sec-debug-the-code}

The key to fixing bugs and errors is pinpointing the cause. There are two main
strategies for doing this:

* Walk forward through the code from the beginning.
* Walk backward through the code from the problem.

Both strategies are valid and powerful.

Walking forward can be time-consuming, because you might have to run a lot of
code before getting to anything related to the problem. On the other hand, it's
conceptually simple (you just go one step at a time, as in
@sec-go-one-step-at-a-time) and might help you uncover other problems as well.

Walking backward is often faster for experienced R users, because warnings and
errors have patterns that you learn to recognize over time.

Consider this code, which is a variation on the example in
@sec-read-warnings-and-errors:

```{r text-addition-error}
#| error: true
num1 = "7"
num2 = 5
sum(num1, num2)
```

The expression on line 3 (the call to `sum`) raises the error, but the
expression on line 1 (which sets `num1`) is the actual source of the problem.
When you debug by walking forward, you can stop after each line of code to
check what's changed.

R provides a built-in tool, called a **debugger**, to help you locate errors.
With the debugger, you can interactively run code one step at a time. This is
similar to what was described in @sec-go-one-step-at-a-time, but can even run
code within loops and function calls one step at a time. You can inspect and
even modify variables as you run the code.

While it's most useful for walking forward from a particular expression, you
can also use the debugger to see where an error was raised, and thus walk
backward (conceptually, without running the code) from there.


### Using the Debugger

The `browser` function pauses the running code and starts the R debugger. To
demonstrate what the debugger can do, consider this code:

```{r, eval = FALSE}
# Run this in an R console.
g = function(x, y) (1 + x) * y

f = function(n) {
  total = 0
  for (i in 1:n) {
    browser()
    total = total + g(i, i)
  }

  total
}

f(11)
```

There are two functions here, `f` and `g`. If you run the code in the R
console, R pauses and opens the debugger as soon as evaluation of `f` reaches
the call to `browser`. You can use the following commands to control the
debugger:

* `n` to run the next line
* `s` to "step into" a call
* `c` to continue running the code
* `Q` to quit the debugger
* `where` to print call stack
* `help` to print debugger help

The debugger also provides several other commands, but these are generally the
most useful.

Notice that you can use the debugger to "step into" a call. This means running
the code in the body of the called function one step at a time, which is
helpful if you suspect the bug originates there.


### Other Debugging Functions

R provides several other built-in functions to help with debugging.

The `debug` function inserts a call to `browser` at the beginning of a
function's body. Use `debug` to debug functions that you can't or don't want to
edit. For example:

```{r, eval = FALSE}
# Run this in an R console.
f = function(x, y) {
  x + y
}

debug(f)
f(5, 5)
```

You can use `undebug` to reverse the effect of `debug`:

```{r, eval = FALSE}
# Run this in an R console.
undebug(f)
f(10, 20)
```

The `debugonce` function places a call to `browser` at the beginning of a
function for the next call only. The idea is that you then don't have to call
`undebug`. For instance:

```{r, eval = FALSE}
# Run this in an R console.
debugonce(f)
f(10, 20)
f(3, 4)
```

The `traceback` function prints the **call stack**, or list of called
functions, for the most recent error. This is quite helpful for walking
backward to the source of an error, because it identifies the call where the
error actually happened (which may be several layers deep in the call stack).

:::{.callout-note}
A printout of the call stack is sometimes also called a stack trace.
:::

Finally, the global option `error` can be used to make R enter the debugger any
time an error occurs. Set the option to `error = recover`:

```{r, eval = FALSE}
options(error = recover)
```

Then try this example:

```{r, eval = FALSE}
# Run this in an R console.
bad_fn = function(x, y) {
  stop("Hi")
  x + y
}

bad_fn(3, 4)
```

## Case Study: Bootstrapping Average Penguin Mass

Let's look at a realistic example to see how debugging works in practice. We'll
use the [Palmer Penguins dataset][penguins], which was collected by [Dr.
Kristen Gorman][gorman] at [Palmer Station, Antarctica][palmer]. The dataset is
available to the public through the palmerpenguins package, which was created
by [Alison Horst][horst].  After installing the package, you can access the dataset with this
code:

```{r setup-palmer-penguins}
library("palmerpenguins")

head(penguins)
```

[penguins]: https://allisonhorst.github.io/palmerpenguins/
[gorman]: https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php
[palmer]: https://pallter.marine.rutgers.edu/
[horst]: https://allisonhorst.com/

:::{.callout-warning}
In recent versions of R, the `penguins` dataset is built-in, but has slightly
different column names. If you want to follow along with this case study, make
sure to use the dataset from the package rather than the built-in dataset.
:::

The dataset consists of measurements for 344 penguins on three islands. The
measurements include flipper length, bill width, sex, and body mass.

Imagine that you're a data analyst working on the dataset and you want to use a
script that was written by a previous member of your lab.

:::{.callout-important}
You can view the script [HERE][penguin-bootstrap].

To download the script, click on the "Download raw file" button at the top
right of the code. Download the script to your computer if you want to try
following along.

[penguin-bootstrap]: https://github.com/ucdavisdatalab/workshop_intermediate_r/blob/main/R/debug_case_study.R
:::

After downloading the script, you can use the `source` function to load the
functions it defines into R:

```{r load-case-study}
source("R/debug_case_study.R")
```

The script provides a function `make_bootstrap_plots` which splits the penguins
into groups according to some categorical features like species and sex, then
plots the uncertainty in the average penguin mass per group. Here's the plot
for mass by species:

```{r first-case-study-plot}
#| message: false
#| warning: false
make_bootstrap_plots(penguins, "species")
```

In the plot, the red dots are measured penguins within each group and the black
boxplots show the variability in 1000 estimates of the group means. The
estimates come from a statistical process called the **bootstrap**. It works by
taking samples from the data we already have (this is called **resampling**),
in order to measure how different the samples are from each other.
Statisticians do this because we often want to know how different our
conclusions would be with completely new data, but we can't go back and collect
the new data.

The `make_bootstrap_plots` function works for multiple categorical features.
Here's how to estimate the means of penguins grouped by species and sex, which
we can use to identify species where one sex is larger than the other:

```{r sex-and-species}
#| message: false
#| warning: false
make_bootstrap_plots(penguins, "species", "sex")
```

Unfortunately, if you group the penguins by species and island, the code raises
an error. This is our bug:

```{r species-island-error}
#| message: false
#| warning: false
#| error: true
make_bootstrap_plots(penguins, "species", "island")
```

### Fixing the Bug

We need to pinpoint the source of the bug, identify the problem, and then fix
it. There are a lot of ways to do this, but they all begin with carefully
reading the error message! In this case, the message is:

```{.none}
Error in `indx[[i]] <- sample(1:N, size = 1)`:
! attempt to select less than one element in integerOneIndex
```

According to the message, there was an attempt to select less than one element
from something. The error message also shows the expression where it was
raised, which appears to take a random sample (the `sample` function) from
a sequence of numbers from 1 to `N`.

At this point, there are many ways to proceed. We could use the `traceback`
function to learn more about where the error was raised, or add `print` calls
to the code to print out details, or walk forward through the code one
expression at a time, either manually or with the debugger. Let's start with
`traceback`:

```{r traceback}
#| eval: false
traceback()
```

```{.none}
4: resample(group$body_mass_g, B = B) at debug_case_study.R#50
3: bootstrap(data, B = B, ...) at debug_case_study.R#60
2: stack(bootstrap(data, B = B, ...)) at debug_case_study.R#60
1: make_bootstrap_plots(penguins, "species", "island")
```

The call stack shows the history of calls leading up to the error, from
innermost (4) to outermost (1). The error is raised in the `resample` function,
which is one of the functions defined by the script. You can inspect the script
to find the line in the body of the `resample` function that matches the
expression mentioned in the error message. It's line 16 in the script.

Remember that the location of an error and the cause of an error are not always
the same thing. To try and understand the cause, let's start the debugger by
putting a `browser` call just before line 16, like this:

```{r resample-with-browser}
#| eval: false
resample = function(data, B) {
  N = length(data)
  result = numeric(B)

  # repeat this whole process B times
  for (b in 1:B) {

    # make an object to hold the resampled indices
    indx = integer(N)

    # resample one data point at a time
    for (i in 1:N) {
      browser()
      indx[[i]] = sample(1:N, size=1)
    }

    # calculate the mean of this resampled data
    result[[b]] = mean(data[indx])
  }

  # return all of the mean resamples
  result
}
```

Now when you run `make_bootstrap_plots(penguins, "species", "island")`, R will
pause and run the debugger at the call to `browser`. You can examine the values
of variables and step forward through the code to see how the values change.

:::{.callout-tip}
In RStudio, you can use the Environment pane to check the values of variables,
and the graphical interface to control the debugger.
:::

Why aren't we seeing an error right away? It's probably because the expression
that raised the error is inside of two for loops. The loops might run for many
iterations before the error occurs. In other words, we need to know which
iteration raises the error, not just which expression in the code.

A simple way to figure this out is to replace the call to `browser` with code
that prints out something to identify the current iteration, such as the values
of `b` and `i`. Then we can run the code until the error occurs, and check the
last thing that was printed. We'll use the `message` function to do this (since
it can paste together several pieces of information):

```{r resample-with-messages}
#| eval: false
resample = function(data, B) {
  N = length(data)
  result = numeric(B)

  # repeat this whole process B times
  for (b in 1:B) {

    # make an object to hold the resampled indices
    indx = integer(N)

    # resample one data point at a time
    for (i in 1:N) {
      message("b=", b, " i=", i)
      indx[[i]] = sample(1:N, size=1)
    }

    # calculate the mean of this resampled data
    result[[b]] = mean(data[indx])
  }

  # return all of the mean resamples
  result
}
```

Now try calling the `make_bootstrap_plots` function again (here we only show
the last few lines of output):

```{r}
#| eval: false
make_bootstrap_plots(penguins, "species", "island")
```

```{.none}
...
b=1000 i=42
b=1000 i=43
b=1000 i=44
b=1 i=1
b=1 i=0
Error in indx[[i]] <- sample(1:N, size = 1) :
  attempt to select less than one element in integerOneIndex
```

Based on the output, the error occurs in the first iteration of the outer loop
(`b=1`). Surprisingly, the inner loop appears to be counting backward: `i=1` is
printed before `i=0`. The value of `i` should never be `0`, since the inner
loop iterates over `1:N`.

Is `i` set to `0` causing the error? We can check by using `0` to index a
vector with `[[`, as in the original line 16:

```{r case-study-reprex}
#| error: true
c(1, 2, 3)[[0L]]
```

That's the same error message, so `i` set to `0` probably is the source of the
problem! But why is `i` getting set to `0`?

Remember what `traceback` told us: `resample` is called from the `bootstrap`
function. Looking where that happens (line 53), we can see that this line is
also inside of a loop. So we have a new call to `resample` in every iteration
of the loop. That explains why there was so much output after we added the call
to `message` to the `resample` function.

We need to figure out which iteration of the loop in the `boostrap` function is
causing a problem. So let's reuse the same strategy: we'll add a call to
`message` to the loop in `bootstrap`, somewhere before line 53, to print out
details about the current iteration:

```{r bootstrap-with-message}
#| eval: false
bootstrap = function(data, ..., B=1000) {
  # get the names of grouping factors
  grouping_factors = sapply(list(...), unlist)

  # identify all possible combinations of grouping factors
  grouping_levels =
    sapply(grouping_factors,
           function(grp) unique(data[[grp]]) |> as.character(),
           simplify = FALSE) |>
    setNames(grouping_factors)

  # cross the grouping levels to get all combinations
  groups = split(data, data[,grouping_factors])

  # create an empty data.frame to hold the results
  col_names = names(groups)
  result = replicate(length(groups), numeric(B), simplify=FALSE) |>
    as.data.frame(col.names = col_names)

  # bootstrap the mean of mass for each group
  for (i in 1:length(groups)) {
    message("group=", i)

    # get the subset of data that is relevant to this group
    group = groups[[i]]

    # calculate the mean of a bunch of bootstrap resamples
    result[[i]] = resample(group$body_mass_g, B=B)
  }

  # return the result
  result
}
```

Now when we call `make_bootstrap_plots`, we get:

```{r}
#| eval: false
make_bootstrap_plots(penguins, "species", "island")
```

```{.none}
...
b=1000 i=42
b=1000 i=43
b=1000 i=44
group=2
b=1 i=1
b=1 i=0
Error in indx[[i]] <- sample(1:N, size = 1) :
  attempt to select less than one element in integerOneIndex
```

It looks like the error happens with the second group. Change the `message`
call to a `browser` call, run the code again, and inspect the second group:

```{.none}
Browse[1]> groups[[1]]
# A tibble: 0 × 8
# ℹ 8 variables: species <fct>, island <fct>, bill_length_mm <dbl>, bill_depth_mm <dbl>,
#   flipper_length_mm <int>, body_mass_g <int>, sex <fct>, year <int>
```

Ahhh, the second group is empty: it has no rows! There were simply no penguns
in the data for some combinations of species and island! That's why `i` gets
set to `0` and the code ends up trying to assign the `0`th entry of `indx`.

Now that we've pinpointed the problem and identified its cause, how can we fix
it? Hint: there are *many* solutions to this bug.

One solution is to avoid resampling empty groups. To achieve this, we need to
replace `1:N`, which returns the vector `c(1, 0)` when `N` is `0` (a common
source of bugs!), with something else. Specifically, we'll use `seq_len(N)`,
which returns an empty vector (with no elements) when `N` is `0`. Let's swap
that in for `1:N` on line 15 of `debug_case_study.R`:

```{r fixed-resample}
resample = function(data, B) {
  N = length(data)
  result = numeric(B)

  # repeat this whole process B times
  for (b in 1:B) {

    # make an object to hold the resampled indices
    indx = integer(N)

    # resample one data point at a time
    for (i in seq_len(N)) {
      indx[[i]] = sample(1:N, size=1)
    }

    # calculate the mean of this resampled data
    result[[b]] = mean(data[indx])
  }

  # return all of the mean resamples
  result
}
```

Then we can try making the plot again:

```{r fixed-example-run}
#| message: false
#| warning: false
make_bootstrap_plots(penguins, "species", "island")
```

Here you can see that the unobserved species/island combinations have been
simply left empty and we no longer have an error. Yay!


## Measuring Performance

How quickly code runs and how much memory it uses can be just as much of an
obstacle to research computing tasks as errors and bugs. This section describes
some of the strategies you can use to estimate or measure the performance
characteristics of code, so that you can identify potential problems and fix
them.


### Estimating Memory Usage

Running out of memory can be extremely frustrating, because it can slow down
your code or prevent it from running at all.

It's useful to know how to estimate how much memory a given data structure will
use so that you can determine whether a programming strategy is feasible before
you even start writing code. The [central processing units][cpu] (CPUs) in most
modern computers are designed to work most efficiently with [64 bits][64-bit]
of data at a time. Consequently, R and other programming languages typically
use 64 bits to store each number (regardless of type). While the data
structures R uses create some additional overhead, you can use this fact to do
[back-of-the-envelope][bote] calculations about how much memory a vector or
matrix of numbers will require.

[cpu]: https://en.wikipedia.org/wiki/Central_processing_unit
[64-bit]: https://en.wikipedia.org/wiki/64-bit_computing
[bote]: https://en.wikipedia.org/wiki/Back-of-the-envelope_calculation

Start by determining how many elements the data structure will contain. Then
multiply by 64 bits and divide by 8 to convert bits to bytes. You can then
repeatedly divide by 1024 to convert to kilobytes, megabytes, gigabytes, or
terabytes. For instance, an vector of 2 million numbers will require
approximately this many megabytes:

```{r}
n = 2000000
n * (64 / 8) / 1024^2
```

You can even write an R function to do these calculations for you! If you're
not sure whether a particular programming strategy is realistic, do the memory
calculations *before* you start writing code. This is a simple way to avoid
strategies that are inefficient.

If you've already written some code and it runs out of memory, the first step
to fixing the problem is identifying the cause. The [lobstr][] package provides
functions to explore how R is using memory.

[lobstr]: https://lobstr.r-lib.org/

You can use the `mem_used` function to get the amount of memory R is currently
using:

```{r}
library("lobstr")
mem_used()
```

Sometimes the culprit isn't your code, but other applications on your computer.
Modern web browsers are especially memory-intensive, and closing yours while
you run code can make a big difference.

If you've determined that your code is the reason R runs out of memory, you can
use the `obj_size` function to get how much memory objects in your code
actually use:

```{r}
obj_size(1)

x = runif(n)
obj_size(x)

obj_size(mtcars)
```

If a specific object created by your code uses a lot of memory, think about
ways you might change the code to avoid creating the object or avoid creating
the entire object at once. For instance, consider whether it's possible to
create part of the object, save that to disk, remove it from memory, and then
create the another part.


### Benchmarking

**Benchmarking** means timing how long code takes to run. Benchmarking is
useful for evaluating different strategies to solve a computational problem and
for understanding how quickly (or slowly) your code runs. When you benchmark
code, it's important to collect and aggregate multiple data points so that your
estimates reflect how the code performs on average.

R has built-in functions for timing code, but several packages provide
functions that are more convenient for benchmarking, because they automatically
run the code multiple times and return summary statistics. The two most mature
packages for benchmarking are:

* [microbenchmark][]
* [bench][]

[microbenchmark]: https://cran.r-project.org/web/packages/microbenchmark/index.html
[bench]: https://bench.r-lib.org/

The microbenchmark package is simpler to use. It provides a single function,
`microbenchmark`, for carrying out benchmarks. The function accepts any number
of expressions to benchmark as arguments. For example, to compare the speed of
`runif` and `rnorm` (as `A` and `B` respectively):

```{r}
library("microbenchmark")
microbenchmark(A = runif(1e5), B = rnorm(1e5))
```

The `microbenchmark` has parameters to control the number of times each
expression runs, the units for the timings, and more. You can find the details
in `?microbenchmark`.


### Profiling


**Profiling** code means collecting data about the code as it runs, and a
**profiler** is a program that profiles code. A typical profiler estimates how
much time is spent on each expression (as actual time or as a percentage of
total runtime) and how much memory the code uses over time. Profiling is a good
way to determine which parts of your code are performance bottlenecks, so that
you can target them when you try to optimize your code.

R has a built-in profiler. You can use the `Rprof` function to enable or
disable the profiler. Essential parameters for the function are:

* `filename` -- a path to a file for storing results. Defaults to `Rprof.out`.
* `interval` -- the time between samples, in seconds.
* `memory.profiling` --  whether to track memory in addition to time.

Set these parameters in the first call you make to `Rprof`, which will enable
the profiler. Then run the code you want to profile. At the end of the code,
call `Rprof(NULL)` to disable the profiler.

The profiler saves the collected data to a file. You can use the `summaryRprof`
function to read the profile data and get a summary. Essential parameters for
this function are:

* `filename` -- the path to the results file. Defaults to `Rprof.out`.
* `memory` -- how to display memory information. Use `"both"` to see total
  changes.

The summary lists times in seconds and memory in bytes.

The [profvis][] package provides an interactive graphical interface for
exploring profile data collected with `Rprof`. Examining profile data
graphically makes it easier to interpret the results and to identify patterns.

[profvis]: https://profvis.r-lib.org/
