Intermediate R Assessment
=========================

_GradPathways Microcredential_

**Instructions:**

Write your solution to each exercise directly beneath the text for THAT
exercise. Put text in comments, but leave code uncommented. Submit the edited
copy of this file.

If you need help, come to DataLab's office hours.


Exercise 1
==========

Consider this code:

```{r, eval = FALSE}
gen = function(x)
{
   function(a) x + a
}
f = gen(1:10)

w = 101
f(w)
```

In the call `f(w)`, where will R find the values of `a` and `x`, and what are
their values?



Exercise 2
==========

The base package contains a variable named `version`. This is used in the
function `is.R`:

```{r}
is.R
```

Suppose you create a variable in your R session (global environment) named
`version`, for example:

```{r}
version = "1.3-0"
```

and then call `is.R()`. Where will `is.R` find the variable `version` when it
evaluates `version$language`?


Exercise 3
==========

Suppose you generate several uniform random values between 0 and 1:

```{r}
x = runif(5)
```

Format these as two single, comma-separated strings:

1. In standard notation with 2 digits after the decimal.
2. In scientific notation with 3 digits after the decimal.


Exercise 4
==========

1. Read (or reread) the [Global Options][global] section in the reader and the
   `?Startup` help file. Pay particular attention to the parts about
   `.Rprofile`. By default, R searches for `.Rprofile` in your home directory.

   Then read `?path.expand` about the `path.expand` function. Use the function
   to locate your home directory.

   [global]: https://ucdavisdatalab.github.io/workshop_intermediate_r/output-errors-and-bugs.html#global-options

2. Create or edit the `.Rprofile` file in your home directory. Make sure the
   file defines a function named `.First`. In that function, use the `options`
   functions to set some options (you can read `?options` to see what's
   possible). At a minimum, set `options(warn = 1)` so that R always prints
   warnings immediately. Restart R and verify that R recognizes your new
   settings. Include a copy of the code in your `.Rprofile` as your answer to
   this part.


Exercise 5
==========

The lubridate function `parse_date_time` is a convenient compromise between the
automatic parsing functions such as `mdy` and the more flexible `fast_strptime`
function. The idea is that you pass `parse_date_time` the order in which the
components of the date or time appear, but don't have to pass an entire date
format string.

1.  Read the documentation for `parse_date_time`. Which parameter controls the
    order in which the date or time components appear? For times and
    date-times, does the function return a `POSIXct` or `POSIXlt` object by
    default?

2.  Use `parse_date_time` to parse these times (make sure to capture AM/PM):

    ```{r}
    x = c("3:15pm", "4 o' clock, 17 minutes past, pm", "08 45 am", "9th hour,
          23th minute, pm")
    ```


Exercise 6
==========

Write a function `parse_numeric` which parses a vector of strings into numbers.
The function should use functions from the stringr package to automatically
remove whitespace, commas, the currency symbols `$`, `€`, `¥`, `£`, and percent
signs `%`.

Test the function on these numbers and also on a vector of numbers you made up
yourself:

```{r}
x = c("€100.21", "88%", "$1,000,000", "3,421.8782", "    ¥2100")
```


Exercise 7
==========

Write a function `clean_names` that cleans up a vector of strings so that they
are all valid R variable names. A valid name must only contain letters,
numbers, underscores `_`, and dots `.`, and must begin with a letter or a dot.
Use functions from the stringr package to replace whitespace between other
characters with a single underscore `_` and remove any other invalid
characters.

Give the function an option `to_lower` that controls whether the letters are
all converted to lowercase, with default `TRUE`.

Test your function on the following strings:

```{r}
x = c("123hello", "%%%YAY%%%", "result1", "Make Great  Coffee", "  date-time ")
```

The output should be:

```
[1] "hello"             "yay"               "result1"
[4] "make_great_coffee" "datetime"
```


Exercise 8
==========

1. Suppose you want to use R to simulate water temperature in a small section
   of a river. At each time step, your simulation will use the result from the
   previous time step to compute a new result. Each result is a 3-dimensional
   array of temperatures, which will be stored in a 4-dimensional array (the
   4th dimension is time). Each time step is 1 second and you want to simulate
   1 minute of temperatures. What kind of loop should you use?

2. Suppose you want to scrape data from a website with information about the TV
   show "I Love Lucy". Each page contains information about a single episode
   and a link to the page for the next episode. You scraper will extract this
   information from each page, and then use the link to process the next page.
   You're not sure exactly how many pages there are. What kind of loop should
   you use?

3. Suppose you're working with a dataset about parking tickets in Davis. Each
   of the 121 observations records one ticket, and includes the approximate
   street address where the ticket was issued. Your friend developed a function
   `geocode_address` that can convert a street address to latitude and
   longitude coordinates, but hasn't figured out how to vectorize it yet. If
   you want to compute coordinates for all of the observations, what kind of
   loop should you use?
